---
title: "Project 1"
author: "Zainab Saeed"
output: html_document
date: "2024-01-19"
runtime: shiny
---
```{r echo=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(ggplot2)
library(plotly)
library(lubridate)
library(scales)
library(zoo) 
library(scales)
library(lubridate)
library(shiny)
library(dplyr)
library(tidyr)
```
# Northwind Data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```


```{r echo=FALSE,message=FALSE,warning=FALSE}
df_orders <- read.csv("/Users/zainab/Downloads/northwind/orders.csv")
df_orders_line_items <-read.csv("/Users/zainab/Downloads/northwind/orderdetails.csv")
df_products <-read.csv("/Users/zainab/Downloads/northwind/products.csv")
df_category <-read.csv("/Users/zainab/Downloads/northwind/categories.csv")
df_employes <-read.csv("/Users/zainab/Downloads/northwind/employees.csv")
df_customer <- read.csv("/Users/zainab/Downloads/northwind/customers.csv")
df_region <- read.csv("/Users/zainab/Downloads/northwind/region.csv")
df_shippers <- read.csv("/Users/zainab/Downloads/northwind/shippers.csv")
df_suppliers <- read.csv("/Users/zainab/Downloads/northwind/suppliers.csv")
df_teritories<- read.csv("/Users/zainab/Downloads/northwind/territories.csv")
df_emp_terr<- read.csv("/Users/zainab/Downloads/northwind/employeeterritories.csv")
```
 
## Category
### What is the average Order Value for each category?
```{r warning=FALSE}
result <- df_orders_line_items  |>
  dplyr::full_join(df_products,
                   by = c("ProductID" = "ProductID")) |>
  dplyr::full_join(df_category,
                   by = c("CategoryID" = "CategoryID")) |>
  dplyr::mutate(subtotal = UnitPrice.x * Quantity * (1 - Discount)) |>
  dplyr::group_by(OrderID, CategoryName) |>
  dplyr::summarise(totalsale = sum(subtotal), .groups = 'drop') |>
  dplyr::ungroup() |>
  dplyr::group_by(CategoryName) |>
  dplyr::summarise(avgsale = mean(totalsale), .groups = 'drop')
print(result)
# Create ggplot object
ggplot_obj <- ggplot(result, aes(x = CategoryName, y = avgsale, fill = CategoryName)) +
  geom_col() +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) +
  theme_minimal()  # You can customize the theme as needed

# Create plotly object and adjust size
plotly_obj <- plotly::ggplotly(ggplot_obj, height = 400, width = 950)

# Show the plotly object
plotly_obj

```

### What is the Quaterly sales revenue for each Category?
```{r}

df_orders  |>
  dplyr::select(OrderID,OrderDate,)|>
  dplyr::full_join(df_orders_line_items|>
                     dplyr::select( OrderID ,ProductID, UnitPrice, Quantity),
                   by = c("OrderID" = "OrderID")) |>
  dplyr::full_join(df_products|>
                     dplyr::select( CategoryID ,ProductID),
                   by = c("ProductID" = "ProductID")) |>
  dplyr::full_join(df_category|>
                     dplyr::select( CategoryID ,CategoryName),
                   by = c("CategoryID" = "CategoryID")) |>
  dplyr::mutate(subtotal = UnitPrice * Quantity,
                Quarter = quarter(as.Date(OrderDate))) |>
  dplyr::group_by(Quarter, CategoryName) |>
  dplyr::summarize(total_sales = sum(subtotal))|>
  plot_ly(x = ~Quarter, y = ~total_sales, type = 'bar', color = ~CategoryName) %>%
  layout(title = "Total Sales by Category and Quarter",
         xaxis = list(title = "Quarter"),
         yaxis = list(title = "Total Sales"),
         barmode = 'stack')
```


## Supplier
### What is the total sales revenue for each Supplier?

```{r warning=FALSE}

supplier_revenue <- df_orders_line_items |>
  dplyr::full_join(df_products,
                   by = c("ProductID" = "ProductID")) |>
  dplyr::full_join(df_suppliers,
                   by = c("SupplierID" = "SupplierID"))|>
  dplyr::mutate(subtotal = UnitPrice.x * Quantity * (1- Discount)) |>
  dplyr::group_by(CompanyName)|>
  dplyr::summarise(totalsale= sum(subtotal ,na.rm = TRUE))
print(supplier_revenue)
ggplot_obje <- ggplot(supplier_revenue, aes(y = CompanyName, x = totalsale, fill = CompanyName)) +
  geom_col() +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) +
  theme_minimal()  # You can customize the theme as needed

# Create plotly object and adjust size
plotly_obje <- plotly::ggplotly(ggplot_obje, height = 400, width = 950)

# Show the plotly object
plotly_obje
```

### What categories of products do different suppliers specialize in?
```{r}
supplier_product_categories <- df_category|>
  dplyr::full_join(df_products,
                   by = c("CategoryID" = "CategoryID")) |>
  dplyr::full_join(df_suppliers,
                   by = c("SupplierID" = "SupplierID"))
supplier_category_counts <- supplier_product_categories %>%
group_by(CompanyName, CategoryName) %>%
summarise(NumProducts = n()) %>%
arrange(CompanyName, desc(NumProducts))

# Print the resulting dataset
print(supplier_category_counts)
interactive_plot <- ggplot(supplier_category_counts, aes(y = CompanyName, x = NumProducts, fill = CategoryName, text = paste("Category: ", CategoryName, "<br>Num Products: ", NumProducts))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Number of Products in Each Category by Supplier",
       x = "Number of Products",
       y = "Supplier Name",
       fill = "Category Name") +
  theme_minimal()

# Convert ggplot to plotly with tooltips
interactive_plot <- ggplotly(interactive_plot, tooltip = "text")

# Show the interactive plot
interactive_plot
```


## Product
### What is the yearly sales revenue for each Product?

```{r warning=FALSE}
product_revenue <- df_orders  |>
  dplyr::select(OrderID, OrderDate) |>
  dplyr::full_join(df_orders_line_items,
                   by = c("OrderID" = "OrderID")) |>
  dplyr::full_join(df_products,
                   by = c("ProductID" = "ProductID")) |>
  dplyr::full_join(df_category,
                   by = c("CategoryID" = "CategoryID")) |>
  dplyr::mutate(subtotal = UnitPrice.x * Quantity * (1 - Discount),
                Year = lubridate::year(as.Date(OrderDate))) |>
  dplyr::group_by(Year, ProductName, .drop = TRUE) |>
  dplyr::summarize(total_sales = sum(subtotal), .groups = 'drop')
print(product_revenue)
ggplot_obj <- ggplot(product_revenue, aes(y = as.factor(Year), x = total_sales, fill = ProductName)) +
  geom_col() +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) +
  theme_minimal()  
plotly_obj <- plotly::ggplotly(ggplot_obj, height = 400, width = 950)

plotly_obj

```
### In terms of sales value, what has been our best selling top 10 product of all time?

```{r warning=FALSE}
product_sales <- df_orders_line_items  |>
  dplyr::full_join(df_products,
                   by = c("ProductID" = "ProductID")) |>
  dplyr::mutate(subtotal = UnitPrice.x * Quantity * (1- Discount)) |>
  dplyr::group_by(ProductName)|>
  dplyr::summarise(totalsale= sum(subtotal ,na.rm = TRUE),
                   total_quantity = sum(Quantity, na.rm = TRUE))|>
  dplyr::arrange(desc(totalsale))
top_10_products <- product_sales[1:10, ]
print(top_10_products)
```

## Freight
### how freight costs vary across different Countries.
```{r warning=FALSE}
country_avg_freight <- df_orders |>
  group_by(ShipCountry) |>
  summarise(AvgFreight = mean(Freight, na.rm = TRUE))

print(country_avg_freight)
```

### Average Freight Costs Across Countries

```{r warning=FALSE}
ggplot_obj <- ggplot(country_avg_freight, aes(x = ShipCountry, y = AvgFreight)) +
  geom_col(fill = "skyblue") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) +
  theme_minimal()  
plotly_obj <- plotly::ggplotly(ggplot_obj, height = 400, width = 1100)

plotly_obj


```


### Calculate the average, minimum, and maximum freight costs
```{r warning=FALSE}
freight_summary <- df_orders |>
  summarise(
    AvgFreight = mean(Freight, na.rm = TRUE),
    MinFreight = min(Freight, na.rm = TRUE),
    MaxFreight = max(Freight, na.rm = TRUE)
  )

# Print the result
print(freight_summary)
```
### Investigate whether freight costs have changed over time. Are there trends or patterns in shipping expenses across different months, or years?
```{r warning=FALSE}
df_orders$OrderDate <- ymd_hms(df_orders$OrderDate)
df_orders$Month <- format(df_orders$OrderDate, "%Y-%m")
df_orders$Year <- format(df_orders$OrderDate, "%Y")

monthly_analysis <- df_orders %>%
  group_by(Month) %>%
  summarise(AvgFreight = mean(Freight, na.rm = TRUE))
print(head(monthly_analysis))
```

### Average Freight Costs Over Months
```{r warning=FALSE}
ggplot(monthly_analysis, aes(x = Month, y = AvgFreight, group=1)) +
  geom_line(color = "red") +
  geom_point(color = "red") +
  labs(title = "Average Freight Costs Over Months",
       x = "Month",
       y = "Average Freight Cost") +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r warning=FALSE}
yearly_analysis <- df_orders %>%
  group_by(Year) %>%
  summarise(AvgFreight = mean(Freight, na.rm = TRUE))
print(head(yearly_analysis))
```

### Average Freight Costs Over Year
```{r warning=FALSE}
ggplot(yearly_analysis, aes(x = Year, y = AvgFreight , group=1)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Average Freight Costs Over Year",
       x = "Year",
       y = "Average Freight Cost") +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Rolling Window
## Employee
### Explore the performance of employees or territories using rolling metrics

```{r}
df_orders$OrderDate <- as.Date(df_orders$OrderDate)

employee_data <- df_employes %>%
  full_join(df_orders, by = c("EmployeeID" = "EmployeeID")) %>%
  full_join(df_orders_line_items, by = c("OrderID" = "OrderID")) %>%
  mutate(subtotal = UnitPrice * Quantity * (1 - Discount))

employee_sales <- employee_data %>%
  select(OrderDate, FirstName, subtotal) %>%
  tidyr::complete(
    OrderDate = seq(from = min(df_orders$OrderDate), to = max(df_orders$OrderDate), by = "days"),
    FirstName = unique(FirstName),
    fill = list(subtotal = 0)
  )

rolling_employee_data <- employee_sales %>%
  arrange(FirstName, OrderDate) %>%
  group_by(FirstName) %>%
  mutate(sales_past_30d = zoo::rollapplyr(subtotal, width = 30, FUN = sum, fill = NA, align = "right", partial = TRUE))

print(rolling_employee_data)
# Your ggplot code with OrderDate included in the text aesthetic
gg <- ggplot(rolling_employee_data, aes(x = OrderDate, y = sales_past_30d, color = FirstName, group = FirstName, text = paste("Order Date: ", OrderDate, "<br>Sales: ", scales::dollar(sales_past_30d, digits = 3)))) +
  geom_line() +
  labs(title = "Rolling Past 30 Days Sales",
       y = "Sales (in dollars)") +
  facet_wrap(~FirstName, scales = "free_y", ncol = 2, drop = FALSE, strip.position = "bottom") +
  theme_minimal() +
  theme(axis.text.x = element_blank())  # Hide x-axis labels

# Convert ggplot to plotly with increased size
interactive_plot <- ggplotly(gg, tooltip = "text", height = 900, width = 900)

# Print or display the interactive plot
interactive_plot

```
## Products
### What are the rolling past 30 days sales trends for different products?
```{r}
# Check for missing values in OrderDate
if (sum(is.na(df_orders$OrderDate)) > 0) {
  stop("Missing values in OrderDate column. Handle them appropriately.")
}

df_result <- df_orders %>%
  full_join(df_orders_line_items %>%
              select(OrderID, ProductID, UnitPrice, Quantity),
            by = "OrderID") %>%
  full_join(df_products %>%
              select(ProductID, ProductName, CategoryID),
            by = "ProductID") %>%
  full_join(df_category %>%
              select(CategoryID, CategoryName),
            by = "CategoryID") %>%
  mutate(subtotal = UnitPrice * Quantity) %>%
  select(OrderDate, ProductName, subtotal) %>%
  tidyr::complete(OrderDate = seq(from = min(df_orders$OrderDate), to = max(df_orders$OrderDate), by = "days"),ProductName = unique(ProductName),fill = list(subtotal = 0))

rool_data <- df_result|>
  arrange(ProductName ,OrderDate)|>
  group_by(ProductName) |>
  mutate(sales_past_30d = zoo::rollapplyr(subtotal, width = 30, FUN = sum, fill = NA, align = "right", partial = TRUE))
print(rool_data)
shinyApp(
  ui = fluidPage(
    titlePanel("Dynamic Line Plot"),
    sidebarLayout(
      sidebarPanel(
        selectInput("product_selector", "Select Products:", choices = unique(rool_data$ProductName))
      ),
      mainPanel(
        plotlyOutput("line_plot")
      )
    )
  ),
  server = function(input, output) {
    # Filter data based on selected employee
    selected_data <- reactive({
      rool_data[rool_data$ProductName == input$product_selector, ]
    })
    
    # Render the dynamic plot
    output$line_plot <- renderPlotly({
      plot_ly(data = selected_data(), x = ~OrderDate, y = ~sales_past_30d,
              text = ~paste("ProductName: ", ProductName, "<br>Date: ", OrderDate, "<br>Sales: ", scales::dollar(sales_past_30d, digits = 3)),
              type = "scatter", mode = "lines",
              width = 1200, height = 600) %>%
        layout(title = paste("Rolling Past 30 Days Sales for", input$product_selector),
               xaxis = list(title = "Order Date"),
               yaxis = list(title = "Sales (in dollars)"),
               showlegend = TRUE,
               hovermode = "closest",
               width=800,
               height= 300)
    })
  }
)

```

### What is the proportion of sales contributed by each city within a country?
```{r message=FALSE,warning=FALSE}
df_total_sale <- df_orders_line_items |>
  dplyr::full_join(df_orders, by = c("OrderID" = "OrderID")) |>
  dplyr::mutate(subtotal = UnitPrice * Quantity * (1 - Discount)) |>
  dplyr::group_by(ShipCity, ShipCountry) |>
  dplyr::summarize(city_sales = sum(subtotal)) |>
  dplyr::group_by(ShipCountry) |>
  dplyr::mutate(country_sales = sum(city_sales)) |>
  dplyr::mutate(p_country_sales = round(city_sales / country_sales, 2))
print(df_total_sale)
# Create ggplot object
gg_plot <- ggplot(df_total_sale, aes(x = ShipCountry, y = p_country_sales, fill = ShipCity)) +
  geom_bar(stat = "identity")

# Convert to plotly
plotly_plot <- plotly::ggplotly(gg_plot)

# Modify layout to increase width
layout(plotly_plot, width = 1200)


```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
